---
title: "Hyla call detection OSU"
output: html_document
date: "`r Sys.Date()`"
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Automatic call detection using R {.tabset .tabset-pills}

### 1) The basics

Before jumping to detecting calls, it's important to understand a few things about amplitude envelopes... For this we will work with a sample of a *Hyla chrysoscelis* call.

First, load the [seewave](https://cran.r-project.org/web/packages/seewave/index.html) and [tuneR](https://cran.r-project.org/web/packages/tuneR/index.html) libraries (make sure you have installed them first with the *install.packages()* function):

```{r warning = FALSE, message = FALSE}
library(seewave)
library(tuneR)
library(scales) #just for the alpha() function to make colors transparent
library(shiny)

# Colorblind friendly colors
blue_col <- "#004488"
red_col <- "#BB5566"
yellow_col <- "#DDAA33"

```


There should be seven audio (wav) files in the working directory. Lets upload the sample call (file: "Hyla_test.wav"):

```{r}
setwd("/Users/matiasvumac/Desktop/Rmarkdown Hyla detector") #define yourself
path_to_wav <- paste(getwd(), "/","Hyla_test.wav", sep = "")
wav <- readWave(path_to_wav) #requires package 'tuneR'.
wav

list.files(path = getwd(), pattern = "\\.wav$") #Check the presence of other WAVs in the wd.
```

Plot the oscillogram using the *oscillo()* function from the library "seewave":

```{r echo=TRUE, fig.align='center'}
oscillo(wav)
```

We can compute the **amplitude envelope** of the oscillogram in R using the function *env()*. Let's see:

```{r, fig.align='center'}
env(wav, 
    envt = "abs", # type of envelope
    colwave = red_col) # color of envelope
```

We can overlay the oscillogram to the envelope:

```{r, fig.align='center'}

oscillo(wav, 
        colwave = alpha("black", 0.3))
par(new = TRUE)
env(wav, 
    envt = "abs",
    colwave = red_col)

```

We can have a closer look at the relationship between oscillogram (black) and amplitude envelope (red) by zooming into a 0.05 seconds section of the call:

```{r, fig.align='center'}

oscillo(wav, 
        colwave = alpha("black", 0.3),
        from = 0.61, 
        to = 0.66) # zoom in the oscillogram to [0.61, 0.66] seconds

par(new = TRUE)
env(wav, 
    envt = "abs",
    colwave = red_col,
    from = 0.61,
    to = 0.66)

```

There are parameters we can manipulate to change how the amplitude envelope is computed - the window size and overlap:

```{r, fig.align='center', fig.dim=c(10, 4)}
wl <- 256  # in samples
ovlp <- 99 # in percent

par(mfrow = c(1, 2), mar = c(4.1, 4.1, 0.28, 0.28)) # Just to create a figure with 1 row and 2 columns
oscillo(wav,
        colwave = alpha("black", 0.3))

par(new = TRUE)
env(wav, 
    envt = "abs", 
    colwave = red_col,
    msmooth = c(wl, ovlp))

#Zoom in
oscillo(wav,
        colwave = alpha("black", 0.3),
        from = 0.61,
        to = 0.66)
par(new = TRUE)
env(wav,
    envt = "abs",
    colwave = red_col,
    msmooth = c(wl, ovlp),
    from = 0.61, 
    to = 0.66) # 256 samples window length

```

Let's try different window lenghts:

```{r, fig.align='center'}
oscillo(wav,
        colwave = alpha("black", 0.3),
        from = 0.61,
        to = 0.66)

par(new = TRUE)
env(wav,
    envt = "abs",
    colwave = red_col, 
    msmooth = c(wl, ovlp),
    from = 0.61,
    to = 0.66) # 256 samples window length

par(new = TRUE)
env(wav,
    envt = "abs",
    colwave = blue_col, 
    msmooth = c(wl/6, ovlp),
    from = 0.61, 
    to = 0.66) # 43 samples window length

par(new = TRUE)
env(wav, 
    envt = "abs",
    colwave = yellow_col,
    msmooth = c(wl*2, ovlp),
    from = 0.61,
    to = 0.66) # 512 samples window 
```

So... for detecting pulses small window lengths works better, and for detecting calls large window lenghts are preferred

```{r, fig.align='center'}

oscillo(wav, 
        colwave = alpha("black", 0.3))

par(new = TRUE)
env(wav, 
    envt = "abs", 
    colwave = blue_col,
    msmooth = c(wl/6, ovlp)) # 43 samples window length

par(new = TRUE)
env(wav,
    envt = "abs", 
    colwave = "darkorange",
    msmooth = c(wl*3, ovlp)) # 768 samples window length
```

A shiny app for testing different combinations of windows lengths and overlaps.

```{r echo=FALSE}
inputPanel(
  selectInput("wl", label = "Number of samples in window:",
              choices = c(4, 8, 16, 32, 64, 128, 256, 512, 1024, 2046), selected = 512),
  
  sliderInput("ovlp", label = "Overlap between windows:",
              min = 0, max = 0.9, value = 0.5, step = 0.1),
  
  radioButtons("envt_type", label = "Choose a type of envelope:",
               choices = c("abs", "hil"), selected = "abs")
)

renderPlot({
  par(mfrow = c(1, 2), mar = c(4.1, 4.1, 0.28, 0.28)) # Just to create a figure with 1 row 
  
  #Full oscillogram
  oscillo(wave = wav, colwave = "grey50")
  par(new = TRUE)
  env(wav, envt = "abs", colwave = red_col, msmooth = c(as.numeric(input$wl), as.numeric(input$ovlp)))
  
  #Zoom in 0.05s
  oscillo(wave = wav, from = 0.61, to = 0.66, colwave = "grey50")
  par(new = TRUE)
  env(wav, from = 0.61, to = 0.66, colwave = red_col, envt = as.character(input$envt_type), msmooth = c(as.numeric(input$wl),   as.numeric(input$ovlp)))
  
  
})
```

### 2) Detecting calls with *timer()*

Now that we know the basics about amplitude envelopes, we can use them to detect calls automatically. The 'timer()' function in the 'seewave' package computes the amplitude envelope of a waveform, and then detects calls using an **user-defined amplitude threshold**. We need to also define the:

1. windows length (in samples)
2. overlap (as a percentage)
3. type of envelope (abs or hil)
4. amplitude threshold (as a percentage)
5. power exponential (numeric)
6. minimum duration of detection (in seconds)


Let's try to detect the calls in the sample call.

```{r, fig.align='center', fig.dim=c(10, 4)}
wl_detect <- 32  # samples
ovlp_detect <- 90 # percentage
thrs_detect <- 5  # percentage
pwr_detect <- 2 # exponential applied to oscillogram
dmin_detect <- 0.004 # minimum duration for detections

detection <- timer(wave = wav,
                   threshold = 5,
                   msmooth = c(wl_detect, ovlp_detect),
                   power = pwr_detect,
                   dmin = dmin_detect,
                   plot = TRUE)
```

Lets see what the output looks like:
```{r}
detection
```
```{r}
library(knitr)
  
# Function to create Raven readable file
  Raven.form <- function(dat.fram)
    {
    dat.fram[ , c("Selection", "View", "Channel", "Low Freq (Hz)", "High Freq (Hz)")] <- c( rep(1: nrow(dat.fram)), rep("Spectrogram 1", nrow(dat.fram)), rep(1, nrow(dat.fram)), rep(0, nrow(dat.fram)), rep(22050, nrow(dat.fram)))
    
    dat.fram <- dat.fram[, c("Selection", "View", "Channel", "Begin Time (s)", "End Time (s)", "Low Freq (Hz)", "High Freq (Hz)")]
    
    dat.fram
    }
  
  #Export selections in Raven format
  df0 <- as.data.frame(cbind("Begin Time (s)" = detection$s.start,
                             "End Time (s)" = detection$s.end))
  df.raven <- Raven.form(df0)

  kable(df.raven)

```

We can export the selections made with *timer()* as a .txt file that can be read in Raven Pro (or Lite).
```{r, echo = TRUE}

   write.table(df.raven, file = file.path(getwd(), "Hyla_test.Table.1.selection.txt"),  quote = F, col.names = T, row.names = F, sep = "\t")

```


```{r}
#library(warbleR)
#library(Rraven)

#sound.files <- "Hyla_test.wav"
#selec <- seq(1:length(detection$s.start))
#start <- detection$s.start
#end <- detection$s.end

#df <- cbind.data.frame(sound.files, selec, start, end)

#df_selec_table <- selection_table(df, path = getwd())

#exp_raven(df_selec_table, file.name = "Hyla_test_Raven", khz.to.hz = TRUE, path = getwd())

```



